Dissertation reading notes

### 2023-07-12 Texfiles/Draft1.tex
spelling
Draft and more notes

Standard economic theory suggests that retirement annuities should be highly prized by individuals as a way to
insure against the risk of late death. However, in developed countries rates of annuitization are far below the
levels that theory predicts. In this paper I test two competing hypothesises for the annuity problem: bequests and
pessimistic life expectancy. Under the coalition government in the UK the law regarding the use of private defined
contribution pensions changed. Individuals were no longer forced to annuitise their pension pots and could access
them in a variety of ways such as lump sum withdrawrels or income drawdown and subsequently the number of annuities
sold in the UK dropped precipitously.

Depending on the reason for the lack of annuitization in the UK, the consumption response of retirees to the pension
reform will differ. If individuals do not annuitise because of pessimistic life expectancy I will show that their
consumption should increase. If, on the other hand, individuals do not annuitise because of a bequest motive, consumption
should not change much as a result of the reform. I will solve lifecycle models for both of these cases and simulate
consumption decisions with, and without, forced annuitisation. I will then use a variety of empirical models to measure
the consumption change in early retirement that resulted from the policy reform. The size and magnitude of this change
will be indicative of the mechanism causing the annuitization problem.

The number of individuals over the age of 65 is expected to increase dramatically in the 21st century.

\subsection{Rough plan}
\begin{itemize}
      \item Intro
      \item Lit review
      \item Models
      \item Empirical
            \begin{enumerate}
                  \item Diff in diff
                  \item RDD
                  \item Matching?
                  \item can I use anything I learnt in panel?
                        In some sense the decision to annuitise is a discrete choice problem so I could use something from there.
            \end{enumerate}
      \item Conclusion
\end{itemize}

>>>>

Standard economic theory suggests that retirement annuities should be highly prized by individuals as a way to
insure against the risk of late death \cite{yaari_65}. However, in developed countries rates of annuitization are far below the
levels that theory predicts. In this paper I test two competing hypotheses for the annuity problem: bequests and
pessimistic life expectancy. Under the coalition government in the UK the law regarding the use of private defined
contribution pensions changed. Individuals were no longer forced to annuitise their pension pots and could access
them in a variety of ways such as lump sum withdrawals or income drawdown and subsequently the number of annuities
sold in the UK dropped precipitously.

Depending on the reason for the lack of annuitization in the UK, the consumption response of retirees to the pension
reform will differ. If individuals do not annuitise because of pessimistic life expectancy I will show that their
consumption should increase. If, on the other hand, individuals do not annuitise because of a bequest motive, consumption
should not change much as a result of the reform. I will solve lifecycle models for both of these cases and simulate
consumption decisions with, and without, forced annuitization. I will then use a variety of empirical models to measure
the consumption change in early retirement that resulted from the policy reform. The size and magnitude of this change
will be indicative of the mechanism causing the annuitization problem.

The importance of retirement policy to individuals in the UK is growing. The number of individuals of
pensionable age is expected to grow from 11.9 million in 2020 to 15.2 million in
2045 according to the latest ONS statistics and for every 1000 people of working age there will be 341 of pensionable
age in 20145 compared to 280 in 2020 \cite{ons_population_predictions_2020}. The increase in absolute and relative
numbers of elderly makes retirement policy more important. Moreover, private, defined contribution (DC), pensions are
becoming increasingly common and are predicted to grow as current cohorts age \cite{cribb_karjalainen_ifs_2023}.
Therefore, policies regarding how private pensions can be accessed will have a larger impact on overall welfare for
retirees.

The so called "pensions freedom act" received Royal Asset in December 2014 marked the end of a series
of pension reforms carried out by the coalition government between 2010 and 2015. The reform was announced
in the Spring budget and made it possible to withdraw money from a private pension pot subject to the
marginal rate of income tax that an individual faced. In the June 2010 budget the government made a first
reform to the annuitization rules, creating an minimum income requirement above which individuals would not need to
annuitise more \cite{finance_act_hmt_2011}. However, this was set at £20,000 and therefore few individuals
were eligible. The minimum income requirement was scrapped in the 2014 bill finally eliminating the
compulsory annuities market. The impact of the reforms on annuity demand has been documented by \cite{cannon_et_al_nier_2016}.
Using data from the Association of British Insurers they show that annuity demand dropped by 75\% from
its maximum.

\subsection{Literature review}
My paper draws on three main strands of literature. The annuity problem, the retirement saving
problem and lifecycle models. \textbf{this intro is rubbish}

\cite{yaari_65} was the first to show that under standard assumptions we would expect individuals to
annuitise all of their wealth at retirement to insure against the risk of long life. Since then there
has been much literature discussing possible reasons that people do not annuitise. \cite{finkelstein_porteba_2002}
and \cite{finkelstein_porteba_2004} find evidence of adverse selection, thereby making the 'money's worth'
of annuities lower for the general population as opposed to the population of annuitants. However, they
also find that theory would still predict annuitization.

\cite{friedman_warshawsky_qje_1990} show that annuitization decisions can be fully explained by a mixture of
bequest motives and actuarily unfair annuities. They solve an augmented life-cycle model with a range of
parameters on how severe the rate of return is on the annuity versus market rates. For plausible values
they find that individuals would optimally not annuitise much wealth. Similarly to \cite{finkelstein_porteba_2004},
\cite{friedman_warshawsky_chicago_1988} show that there is a significant difference between the life expectancy of
annuitants and the general population in the American annuity market but this cannot fully explain the annuitization
problem. Only when bequest motives are added to the model can annuitization rates be rationalized.

\cite{lockwood_red_2012} builds on this and shows that a realistic bequest motive in lifecycle simulations achieves
realistic annuitization rates. He solves a simple lifecycle model with bequest motives taken from several recent papers
in the literature. The bequest motives he picks therefore match other important aspects of the lifecycle model such
as how much individuals actually bequest and how rich individuals are when they bequest.

\cite{mitchell_et_al_1999} - this is empirical
\cite{friedman_warshawsky_qje_1990} - this is theory

\cite{friedman_warshawsky_chicago_1988} - this is more theory. Probably good to have read.

\cite{lockwood_aer_2018}

. \textit{Continue
      chat about how elderly population is growing and private pension pots are growing in size, could be worth referencing
      latest census data so we seem super up to date, look at IFS data about the growth of DC pensions as well. }
\textbf{I will do this this afternoon}

\subsection{Rough plan}
\begin{itemize}
      \item Intro
            \begin{enumerate}
                  \item I think Eric French wrote something about population that I could use in then intro to say why it is important.
                  \item "Latest data from HM Revenue Customs (published in April) showed more than £45bn has been taken from pots since 2015."
                        https://www.ftadviser.com/pensions/2022/02/08/pension-freedoms-were-they-really-a-good-idea/

                  \item Need annuity theory references -- Yarri?
            \end{enumerate}
      \item Lit review
      \item Models
      \item Empirical
            \begin{enumerate}
                  \item Diff in diff
                  \item RDD
                  \item Matching?
                  \item can I use anything I learnt in panel?
                        In some sense the decision to annuitise is a discrete choice problem so I could use something from there.
            \end{enumerate}
      \item Conclusion
\end{itemize}

### 2023-07-12 Texfiles/Notes.tex
Draft and more notes

\section{Personal notes}
\subsection{ELSA notes}

>>>>

\subsection{Government policy timeline}
\begin{itemize}
      \item Freedom and choice in pensions announced in March \cite{pen_freedoms_hmt_2014}. Gains royal
            assent in December becoming the Taxation of Pensions Act. \cite{tax_pensions_hmg_2014}

      \item What other reforms were there? 2010 budget annuitisation
\end{itemize}

\section{The Cost of Annuities: Implications for Saving Behavior and Bequests}
\cite{friedman_warshawsky_qje_1990}

\begin{itemize}
      \item Interesting point re annuities being low because of state pension.
            I could test consumption response to forced annuitization if that is the
            case as well. I think it would just mean increased spending?

      \item Make a good point that other explanations for the AP are to do with unfairly priced
            annuities. This could be subjectively or not and it would have the same impact (or
            a mixture of the two). So I don't think there is massive need in estimating subjective
            survival curves.

      \item I think there has been some work by cannon and tokes on estimating this in the
            early 2010s so I am not sure whether there is point me trying to do this again using the
            CMI annuitants tables.

      \item Also test model with both bequests and unfair annuities. This seems most obvious tbh.
            Maybe whole idea is rubbish because of this.

      \item Could I compare how a mixture of load factors vs a mixture of subjective death probs
            vs a mixture of adverse selection interacts. Disentangling the extent to which each causes
            people not to annuitise would be difficult though. But can use evidence from ELSA and CMI for
            subjective and adverse selection amounts respectively.

      \item \begin{equation*}
                  \mathbb{E}(U) = \sum_{0}^{w - x - 1}[p_{t}U_{t}(c_{t}) + p_{t}q_{t}V_{t+1}(G_{t+1})]
            \end{equation*}
            Where $V_{t+1}(G_{t+1})$ is the utility from bequests next period and $q_{t}$ is the probability of death.

      \item Does some simulations showing how bequests and acturarial unfairness interact with each other.
            This type of table could be a good thing to include in my paper. Then a section showing implied consumption
            changes and then the data work. I think I am kinda close to being able to make that anyway.

      \item What is expected yield differential? D
      \item \textit{" but investigating it seriously would require data on the
                  degree of actuarial unfairness in the pricing of annuities not just for
                  individuals but for double or even multiple lives."} This is interesting.
            Could I compare annuitization decisions of couples and singles? Singles
            should be more likely to buy annuities if this is true.

      \item Couples should also have a larger consumption response. Can we solve
            two person models? Probably? But I think joint annuities would face
            more than double the problem with actuarially unfair annuities worse --
            impact would be multiplied and not summed.

      \item Have not seen this before? Could be interesting. I think Eric solved a life-cycle model
            with a couple in so I could try and look at that paper. (No I think it was another paper from S130)

\section{Annuity Prices and Saving Behavior in the United States}
\begin{itemize}
      \item Solve life cycle models again. The variable of interest they use is the
            loading factor.

      \item Need bequest motives to get annuity rates that are observed.
      \item "See \cite{friedman_warshawsky_nber_1985} for a brief discussion of several
            of these other possible explanations." that is for other explanations for
            lack of annuitization. Had a look at this and it wasn't super useful. Just
            seemed like a first draft of the paper.

\section{Bequest motives and the annuity puzzle}
\begin{itemize}
      \item Literature (Abel 2004?? Find when landed) claims that bequest motives
            can only explain part of the AP. But Lockwood shows that it can explain why people
            annuitise no wealth given loading factors (prices etc) of annuities.

      \item  (Davidoff et al., 2005??) is another paper on annuities I should read.

\section{Personal notes}

\subsection{ELSA notes}

\subsection{\textbf{12-07}}
\begin{itemize}

      \item Got CMI access.
\end{itemize}

\subsection{\textbf{11-07}}
\begin{itemize}
      \item Good idea about couples. Good idea about table showing \% of annuititants
            with param values and then also consumption changes.

      \item How to split consumption? Guess have more grids? -- Don't need to since only one
            continuation value and there is no heterogeneity between partners.
      \item How do behaviourel models change the DP problem. Could present bias
            be part of the explanation.

      \item Also introduction of lifetime cap in care costs could be interesting
            hypothetical policy issue to mention.

      \item
\end{itemize}

\subsection{\textbf{08-07}}
\begin{itemize}
      \item Trying to understand the weibull code from \cite{odea_sturrock_rest_2023}. Non-linear
            least squares
      \item Big docs have questions in that we can use.
\end{itemize}
